---
title: "HW2"
author: "Mikhail Filippov"
date: "2024-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(vegan)
library(gridExtra)
library(ape)
library(dendextend)
library(pvclust)
library(tidyr)
library(golubEsets)
```

# Loading dataset

```{r}
data(Golub_Merge)
golub <- data.frame(Golub_Merge)[1:7129]
rownames(golub) <- paste(rownames(golub), Golub_Merge$ALL.AML, Golub_Merge$BM.PB, Golub_Merge$T.B.cell, sep = '_')
```


# Calculating distances
```{r}

euc_dist <-  vegdist(golub, method = "euclidean")
canb_dist <- vegdist(golub, method = "canberra")
manh_dist <- vegdist(golub, method = "manhattan")
gow_dist<- vegdist(golub, method = "gower")

```

# Evaluating clustering
```{r}
compare_clustering_methods <- function(distances, methods) {
  corr_df <- data.frame(
    Method = character(0),
    Distance = character(0),
    Corr = numeric(0)
  )
  
  for  (method in methods){
    for (dist_name in names(distances)){
      d <- distances[[dist_name]]
      hs <- hclust(d, method = method)
      cophenetic_matrix <- cophenetic(hs)
      correlation <- cor(d, as.dist(cophenetic_matrix))
      corr_df <- rbind(corr_df, data_frame(
        Method = method,
        Distance = dist_name,
        Correlation = correlation
      ))
    }
    
  }
  return(corr_df)
}
```

```{r}
distances <- list(Euclidean = euc_dist, Canberra = canb_dist, Mahattan = manh_dist, Gower = gow_dist)
methods = c('single', 'complete', 'average', 'ward.D2')

comparison_results <- spread(compare_clustering_methods(distances, methods), Distance, Correlation)

comparison_results[['Mean']] <- rowMeans(comparison_results[, 2:4])
comparison_results
```
```{r}
colMeans(comparison_results[, 2:5])
```
Best method is average clustering (UPGMA), and it worked good with Gower and euclidean distances. Second best is single (nearest neighbour). 


# Bootstrapping
## Average clustering
```{r}
cl_boot <- pvclust(
  t(golub),
  method.hclust = 'average',
  nboot = 100,
  method.dist = 'euclidean',
  parallel = T,
  iseed = 13)
```
```{r}
plot(cl_boot, cex = 0.7)
```
## Nearest neighbour clustering

```{r}
cl_boot <- pvclust(
  t(golub),
  method.hclust = 'single',
  nboot = 100,
  method.dist = 'euclidean',
  parallel = T,
  iseed = 13)
```

```{r}
plot(cl_boot, cex = 0.7)
```

# Conclusion

UPGMA clustering appeared to be the most meaningful, as there are distinct clusters by treatment and cell types. Nearest neighbor, on the other hand, clustered all samples in a single large cluster, which is impossible to interpret biologically.
